An investigation into the contentious phylogenetic results of the *Mixta* species

### 2021-01-19 Manuscript/Manuscript_Paper.Rmd
Update Manuscript_Paper.Rmd

---
title: "Untitled"
author: "Kimberly Hinz"
date: "2020/12/08"
output:
  word_document:
    reference_docx: Styles1.docx
    fig_caption: true
    # toc: true
    # toc_depth: 5
bibliography: References.bib
csl: IJSEM.csl
---
```{r Packages, echo = FALSE, message = FALSE, warning=FALSE}
library("knitr")
library("tidyverse")
library("ggplot2")
library("RColorBrewer")
library("extrafont")
theme_set(theme_classic())
```
\newpage
# Abstract
*Mixta* is a recently described genus within the bacterial family *Erwiniaceae* that is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results, and identified genes that may have resulted from horizontal gene transfer. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned With the ClustalW algorithm and MEGA-CC was used for model testing and computing phylogenetic trees for each gene for both nucleotide and amino acid sequences. Distance matrices were then extracted from these trees. Nucleotide and amino acid identity analyses were also done in `R` programming language. *Pantoea* was the closest relative to the *Mixta* species in most of the analyses; however, results were not consistent across analyses. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to the selection of genes for phylogenetic analysis and to the method chosen to prevent xenologous signal from distorting the true relationships. Furthermore, future research should consider that different phylogenetic analyses may give different results.

\newpage
# Introduction
The bacterial family Erwiniaceae includes diverse Gram-negative [@Winslow1917; @Prakash2015] organisms including the genera *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Buchnera*, *Wigglesworthia*, *Phaseolibacter*, *Kalamiella* (LPSN, http://www.bacterio.net/). Three of these, *Mixta*, *Pantoea*, and *Erwinia*, have been the focus of numerous studies given their relevance to plant and human pathogens. For example, the genus *Erwinia* contains plant pathogens (e.g. E. amylovora) [@HaubenSwings2005; @Lopez2011; @Matsuura2012], epiphytes (e.g. E. tasmaniensis) [@Geider2006; @Mergaert1999], species that may contribute to plant disease [@Buonaurio2015], and others that are insect-associated [@Campillo2015; @Harada1997; @Skrodenyte-Arbaciauskiene2012]. Similarly, *Pantoea*, a genus reported by Gavini et al.[@Gavini1989], includes species that inhabit a variety of environments [@Gavini1989; @Popp2010; @Liu2013], including human and plant pathogens [@Prakash2015; @deBaere2004; @Fullerton2007; @Kratz2003; @Lim2006; @Schmid2003; @vanRostenberghe2006] as well as species associated with plant growth [@Smits2011; @Kim2012], insects [@Palmer2016], and fungi [@Ma2016; @Rong2016]. *Mixta*, originally *Pantoea*, is a novel genus first suggested by Palmer et al. [@Palmer2017] and then later proposed by the same researchers in 2018 [@Palmer2018]. Two species, M. calida and M. gaviniae (previously species of *Pantoea*), were originally found in infant formula and its production environment [@Popp2010]. Fritz et al. [@Fritz2014] demonstrated that some strains of M. calida were linked with post-surgical meningitis in humans. Currently, there are only five *Mixta* species validly published (LPSN, http://www.bacterio.net/).

Understanding the phylogeny of bacterial species and genes allows researchers to investigate the evolutionary history of the species, their relatedness, and the individual genes that allow the species to survive under various conditions [@Palmer2017; @Bennett2012; @Prasanna2013; @Angus2014; @deMaayer2014; @Fouts2016]. The most common methods in phylogenetic analysis are those including 16S ribosomal RNA (rRNA) sequences [@Konstantinidis2007], multi-locus sequence analysis (MLSA) with several housekeeping genes [@Konstantinidis2007; @Gevers2005; @Glaeser2015], and ribosomal MLSA with 53 structural ribosomal proteins [@Palmer2017; @Bennett2012; @Jolley2012]. However, the phylogenetic trees generated from these methods may not be very robust [@Konstantinidis2007; @Glaeser2015; @Brady2008] because of the lower phylogenetic signal from the highly conserved housekeeping genes [@Konstantinidis2007; @Gevers2005; @Fox1992; @Konstantinidis2005; @Staley2006; @RichterRosselloMora2009]. This is particularly seen within Erwiniaceae wherein most of our understanding of the phylogeny and interrelationships are based on the 16S rRNA genes [@Francino2006; @Hauben1998; @Naum2008; @Sproer1999; @Adeolu2016] because of its ubiquity. However, it has insufficient resolution at the species level [@Glaeser2015] and the genera *Pantoea*, *Erwinia*, *Tatumella*, and *Mixta* are often not monophyletic [@Rezzonico2016]. Another issue that may arise with MLSA is unknowingly including paralogues (i.e. genes originating from an intragenomic duplication event) or xenologues (i.e. genes originating from horizontal gene transfer) [@Palmer2017; @Koonin2005].

The DNA topoisomerase (gyrase) subunit (gyrB), DNA-directed RNA polymerase subunit (rpoB), ATP synthase subunit (atpD), and initiation translation factor (infB) have been shown to be more reliable than 16S rRNA gene sequences for determining intra- and intergeneric relationships [@Hedegaard1999; @Dauga2002; @Mollet1997; @Paradis2005] and a robust method for the Erwiniaceae family [@Brady2010]. The genera within Erwiniaceae are monophyletic when concatenated partial gyrB, rpoB, atpD, and infB gene sequences are used [@Brady2008; @Brady2009; @Moretti2011].

Nevertheless, despite efforts to find a suitable method for identifying the species tree for Erwiniaceae genera, *Mixta*’s position with relation to *Pantoea*, *Erwinia*, and *Tatumella* vastly differs between studies [@Prakash2015; @Palmer2017; @Palmer2018; @Adeolu2016; @Brady2011; @Rezzonico2016; @Brady2012; @Gueule2015]. Not only do different gene combinations provide different results, but different algorithms and models for alignments and phylogenetic trees also contribute to the contentious results [@Naum2008; @Octavia2014]. For example, Prakash et al. [@Prakash2015], Palmer et al. [@Palmer2017], Rezzonico et al. [@Rezzonico2016], and Gueule et al. [@Gueule2015] all used gyrB, rpoB, atpD, and infB nucleotide gene sequences in their phylogenetic analyses. Prakash et al. conducted a neighbour-joining statistical method while the others used a maximum-likelihood (ML) method. Three of the studies – Rezzonico et al., Prakash et al., and Gueule et al. – placed the *Mixta* species within the *Pantoea* genera whereas the analysis in Palmer et al. placed *Mixta* outside of the other three genera. Palmer et al. [@Palmer2018], Rezzonico et al. [@Rezzonico2016], and Brady et al. [@Brady2012] also ran analyses with the four MLSA genes as amino acid sequences. These three studies show contentious results as Rezzonico et al. found that *Mixta* is related closest with *Tatumella*, Palmer et al. found that *Mixta* is related closest with *Pantoea*, and Brady et al. found *Mixta* within the *Pantoea* genus. In a separate study conducted by Palmer et al. @Palmer2017, two ML trees were computed, one with 1039 core amino acid sequence and the other with 52 ribosomal MLSA nucleotide genes. Both ML trees contained species from *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella*; however, the analysis with the core genes placed *Pantoea* and *Tatumella* as sister taxa followed by two *Mixta* species whereas the analysis with ribosomal MLSA genes placed *Pantoea* and *Tatumella* as sister taxa followed by *Erwinia* and then by the *Mixta* species.

There are two possible causes behind contentious results in phylogenetic analyses. One is the use of different methods, including the use of a different phylogenetic tree model from another study despite both studies conducting maximum likelihood trees and the same sequence type. Additionally, there is no common phylogenetic analysis model between nucleotide and amino acid sequences; therefore, evolutionary history is estimated differently. The other reason may be horizontal gene transfer (HGT) events within the *Erwiniaceae* or from a non-closely-related species. HGT events occur frequently within bacteria and often the genes provide the bacteria the ability to adapt to new environments and/or survive selective pressures.

The objective of this paper was to determine the cause behind the contentious phylogenetic results for the *Mixta* species. To accomplish this, homologous genes were extracted from ten complete or partial genomes - two for each of *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella*, and two outgroups - that were retrieved from public databases. In this study, both the nucleotide and amino acid sequences for the same genes were separately analyzed. Model testing was done for each gene in the dataset and phylogenetic trees were computed using the recommended model. Distance matrices were exported from the phylogenetic trees and these were used to determine which species were most closely related to *Mixta* for each gene. Nucleotide and amino acid identity analyses were also conducted for each gene in order to compare the results between evolutionary history and sequence similarity. Phylogenetic trees were then done with either one gene or multiple genes to confirm results.

# Materials and Methods
## Genomes and Gene Sets
Publicly available full genomes were retrieved from NCBI comprising of ten type strains from *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Enterobacter*, and *Pseudomonas* (Table 1). *Enterobacter cloacae* is in the same order of *Enterobacterales* as *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella* whereas *Pseudomonas syringiae* is in the same class of *Gammaproteobacteria*. The genomes were annotated with Prokka v1.14.1 [@Seemann2014] using default parameters and homologous genes were extracted using GET_HOMOLOGUES software package [@Contreras-Moreira2013] with the bidirectional best-hit search algorithm using default parameters. GET_HOMOLOGUES extracted 954 homologous genes between the ten species.

@@

## Genetic Distance
### Closest Relative
Genetic distance gives a numeric estimate of evolutionary change between species. The species that has the shortest genetic distance to a species of interest is considered to be closest relative. For the following analysis, model testing was done first to estimate the best model for each gene. Next, individual phylogenetic gene trees were created and it was from these that distance matrices were exported. *P. septica* was the closest relative to 66.3% and 66.5% of the *M. calida* and *M. gaviniae* genes, respectively, for the nucleotide sequences (Figure 1). In comparison, only 42.7% and 42.8% of *M. calida* and *M. gaviniae* genes, respectively, are closely related to *P. septica* for the amino acid sequences (Figure 1), but this is still the majority. These results suggest that *P. septica* is the closest relative to *Mixta* followed by *P. agglomerans*. For the amino acid sequences, instead of the *Mixta* genes being more closely related to *P. septica*, they are more closely related to the other species in this study (Figure 1). For example, the percentage of genes related to *P. agglomerans* increased from 13.9% for both of the *Mixta* species to 22.3 and 22.4% to *M. calida* and *M. gaviniae*, respectively. This may be due to a loss of phylogenetic signal when using amino acid sequences as a result of codon degeneracy and having a third of available sites.

@@

If genetic distance infers the same closest relative for a *Mixta* gene for both nucleotide and amino acid analyses, then these results may be more reliable. Thus, for genes where the closest relative is an *Erwinia*, *Tatumella*, or *Enterobacter* species both nucleotide and amino acid analyses, they may have resulted from a horizontal gene transfer (HGT) event rather than vertical evolution from a *Pantoea* species. No *Mixta* genes were found to be closely related to *P. syringiae* for either the nucleotide or amino acid sequences (Figure 1). This confirms that *P. syringiae* is a suitable outgroup for this analysis and suggests that any HGT events may be restricted to the order *Enterobacterales*.

@@

```{r Figure3, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 3.", fig.width=6.5, fig.height=5}
FR_nucl_calida <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure3_Circular_NT.csv", stringsAsFactors = FALSE) %>%
  mutate(Closest_Relative = factor(Closest_Relative, 
                                   levels = c("P_septica", "P_agglomerans", "E_tasmaniensis", "E_amylovora", "T_ptyseos", 
                                              "T_saanichensis", "E_cloacae", "P_syringiae"), ordered = TRUE))

ggplot(FR_nucl_calida, aes(x = ID, y = Relative_Number, fill = Closest_Relative)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 15) +
  coord_polar() +
  scale_fill_brewer(palette = "Paired", 
                    labels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", "E. cloacae",
                               "P. syringiae")) +
  scale_y_continuous(limits = c(0, 9), breaks = c(0, 2, 4, 6, 8)) +
  labs(fill = "Closest Relative") +
  theme_bw() +
  theme(legend.text = element_text(face = "italic"),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(size = 12,  family = "Times New Roman"))
```

@@

```{r Figure4, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 4.", fig.width=6.5, fig.height=5}
FR_aa_calida <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure4_Circular_AA.csv", stringsAsFactors = FALSE) %>%
  mutate(Closest_Relative = factor(Closest_Relative, 
                                   levels = c("P_septica", "P_agglomerans", "E_tasmaniensis", "E_amylovora", "T_ptyseos", 
                                              "T_saanichensis", "E_cloacae", "P_syringiae"), ordered = TRUE))

ggplot(FR_aa_calida, aes(x = ID, y = Relative_Number, fill = Closest_Relative)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 15) +
  coord_polar() +
  scale_fill_brewer(palette = "Paired", 
                    labels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", "E. cloacae",
                               "P. syringiae")) +
  scale_y_continuous(limits = c(0, 9), breaks = c(0, 2, 4, 6, 8)) +
  labs(fill = "Closest Relative") +
  theme_bw() +
  theme(legend.text = element_text(face = "italic"),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(size = 12,  family = "Times New Roman"))
```

@@

## Nucleotide and Amino Acid Identity
### Nucleotide Identity

### Amino Acid Identity

## A Comparison Between Evolutionary Change and Site Identity

## Phylogenetic Trees

Nucleotides: ttuB where M. calida is with T. ptyseos and M. gaviniae is with P. septica
Amino acids: ttuB where M. calida is with T. saanichensis and M. gaviniae is with P. septica 
             secY where M. calida is with E. tasmaniensis and M. gaviniae is with P. agglomerans
# Conclusion

>>>>

---
title: "An investigation into the contentious phylogenetic results of the *Mixta* species"
author: "Kimberly Hinz"
date: "2020/12/08"
output:
  word_document:
    reference_docx: Styles1.docx
    fig_caption: true
    # toc: true
    # toc_depth: 5
bibliography: References.bib
csl: IJSEM.csl
---
```{r Packages, echo = FALSE, message = FALSE, warning=FALSE}
library("knitr")
library("tidyverse")
library("ggplot2")
library("RColorBrewer")
library("extrafont")
theme_set(theme_classic())
```
\newpage
# Abstract
*Mixta* is a recently described genus within the bacterial family *Erwiniaceae* that is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results and identified genes that may have resulted from horizontal gene transfer. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned with the ClustalW algorithm, and MEGA-CC was used for model testing and computing phylogenetic trees for each gene for both nucleotide and amino acid sequences. Distance matrices were then extracted from these trees. Nucleotide and amino acid identity analyses were also done using the programming language `R`. *Pantoea* was the closest relative to the *Mixta* species in most of the analyses; however, results were not consistent across analyses. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to selecting genes for phylogenetic analysis and the method chosen to prevent any xenologous signal from distorting the actual relationships. Furthermore, future research should consider that different phylogenetic analyses may provide different results.

\newpage
# Introduction
The bacterial family Erwiniaceae includes diverse Gram-negative [@Winslow1917; @Prakash2015] organisms, including the genera *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Buchnera*, *Wigglesworthia*, *Phaseolibacter*, *Kalamiella* (LPSN, http://www.bacterio.net/). Numerous studies have focused on three of these genera – *Mixta*, *Pantoea*, and *Erwinia* – given their relevance to plant and human pathogens. For example, the genus *Erwinia* contains plant pathogens (e.g. *E. amylovora*) [@HaubenSwings2005; @Lopez2011; @Matsuura2012], epiphytes (e.g. *E. tasmaniensis*) [@Geider2006; @Mergaert1999], species that may contribute to plant disease [@Buonaurio2015], and others that are insect-associated [@Campillo2015; @Harada1997; @Skrodenyte-Arbaciauskiene2012]. Similarly, *Pantoea*, a genus reported by Gavini et al.[@Gavini1989], includes species that inhabit a variety of environments [@Gavini1989; @Popp2010; @Liu2013], including human and plant pathogens [@Prakash2015; @deBaere2004; @Fullerton2007; @Kratz2003; @Lim2006; @Schmid2003; @vanRostenberghe2006] as well as species associated with plant growth [@Smits2011; @Kim2012], insects [@Palmer2016], and fungi [@Ma2016; @Rong2016]. *Mixta*, originally *Pantoea*, is a novel genus first suggested by Palmer et al. [@Palmer2017] and later proposed by the same researchers in 2018 [@Palmer2018]. *M. calida* and *M. gaviniae* (previously species of *Pantoea*) were initially found in infant formula and its production environment [@Popp2010]. Fritz et al. [@Fritz2014] demonstrated that some strains of *M. calida* were linked with post-surgical meningitis in humans. Currently, there are only five *Mixta* species validly published (LPSN, http://www.bacterio.net/).

Understanding the phylogeny of bacterial species and genes allows researchers to investigate the species' evolutionary histories, relatedness, and the individual genes that enable the species to survive under various conditions [@Palmer2017; @Bennett2012; @Prasanna2013; @Angus2014; @deMaayer2014; @Fouts2016]. The most common methods in phylogenetic analysis are those that use ubiquitous genes including 16S ribosomal RNA (rRNA) sequences [@Konstantinidis2007], multilocus sequence analysis (MLSA) with several housekeeping genes [@Konstantinidis2007; @Gevers2005; @Glaeser2015], and ribosomal MLSA with 53 structural ribosomal proteins [@Palmer2017; @Bennett2012; @Jolley2012]. However, the phylogenetic trees generated from these methods may not be robust [@Konstantinidis2007; @Glaeser2015; @Brady2008] because of the lower phylogenetic signal from the highly conserved genes [@Konstantinidis2007; @Gevers2005; @Fox1992; @Konstantinidis2005; @Staley2006; @RichterRosselloMora2009]. This issue is particularly the case with *Erwiniaceae* because much of our understanding is based on the 16S rRNA and other housekeeping genes [@Francino2006; @Hauben1998; @Naum2008; @Sproer1999; @Adeolu2016]. However, such analyses have insufficient resolution at the species level [@Glaeser2015], and the genera *Pantoea*, *Erwinia*, *Tatumella*, and *Mixta* are often not monophyletic [@Rezzonico2016]. Another issue is the unintentional inclusion of paralogues (originating from an intragenomic duplication event) or xenologues (arising from horizontal gene transfer) in MLSA gene analyses [@Palmer2017; @Koonin2005].

The DNA gyrase subunit (gyrB), DNA-directed RNA polymerase subunit (rpoB), ATP synthase subunit (atpD), and initiation translation factor (infB) are more reliable than 16S rRNA gene sequences for determining intra- and intergeneric relationships [@Hedegaard1999; @Dauga2002; @Mollet1997; @Paradis2005] and a robust method for the Erwiniaceae family [@Brady2010]. The genera within Erwiniaceae are monophyletic when the analysis uses concatenated partial gyrB, rpoB, atpD, and infB gene sequences [@Brady2008; @Brady2009; @Moretti2011].

Nevertheless, despite efforts to find a suitable method for identifying the species tree for Erwiniaceae genera, *Mixta*’s position relative to *Pantoea*, *Erwinia*, and *Tatumella* vastly differs between studies [@Prakash2015; @Palmer2017; @Palmer2018; @Adeolu2016; @Brady2011; @Rezzonico2016; @Brady2012; @Gueule2015]. Not only do different gene combinations provide different results, but different algorithms and models for alignments and phylogenetic trees also contribute to the contentious results [@Naum2008; @Octavia2014]. For example, Prakash et al. [@Prakash2015], Palmer et al. [@Palmer2017], Rezzonico et al. [@Rezzonico2016], and Gueule et al. [@Gueule2015] all used gyrB, rpoB, atpD, and infB nucleotide gene sequences in their phylogenetic analyses. Prakash et al. conducted a neighbour-joining statistical method while the others used a maximum-likelihood (ML) method. Three of the studies – Rezzonico et al., Prakash et al., and Gueule et al. – placed the *Mixta* species within the *Pantoea* genera. In contrast. the analysis in Palmer et al. identified *Mixta* outside of the other three genera. Palmer et al. [@Palmer2018], Rezzonico et al. [@Rezzonico2016], and Brady et al. [@Brady2012] also ran analyses with the four MLSA genes as amino acid sequences.

These three studies showed contentious results as *Mixta* was said to be related closest with *Tatumella* by Rezzonico et al., with *Pantoea* by Palmer et al., and within the *Pantoea* clade by

These three studies showed contentious results as

These three studies show contentious results as Rezzonico et al. found that *Mixta* is related closest with *Tatumella*, Palmer et al. found that *Mixta* is related closest with *Pantoea*, and Brady et al. found *Mixta* within the *Pantoea* genus.

In a separate study conducted by Palmer et al. @Palmer2017, two ML trees were computed, one with 1039 core amino acid sequence and the other with 52 ribosomal MLSA nucleotide genes. Both ML trees contained species from *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella*; however, the analysis with the core genes placed *Pantoea* and *Tatumella* as sister taxa followed by two *Mixta* species whereas the analysis with ribosomal MLSA genes placed *Pantoea* and *Tatumella* as sister taxa followed by *Erwinia* and then by the *Mixta* species.

There are two possible causes behind contentious results in phylogenetic analyses. One is the use of different methods, including the use of a different phylogenetic tree model from another study despite both studies conducting maximum likelihood trees and the same sequence type. Additionally, there is no common phylogenetic analysis model between nucleotide and amino acid sequences; therefore, evolutionary history is estimated differently. The other reason may be horizontal gene transfer (HGT) events within the *Erwiniaceae* or from a non-closely-related species. HGT events frequently occur within bacteria, and often the genes provide the bacteria with the ability to adapt to new environments and/or survive selective pressures.

The objective of this paper was to determine the cause behind the contentious phylogenetic results for the *Mixta* species. Homologous genes were extracted from ten complete or partial genomes - two for each of *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella*, and two outgroups - that were retrieved from public databases. In this study, both the nucleotide and amino acid sequences for the same genes were separately analyzed. Model testing was done for each gene in the dataset, and phylogenetic trees were computed using the recommended model. Distance matrices were exported from the phylogenetic trees, and these were used to determine which species were most closely related to *Mixta* for each gene. Nucleotide and amino acid identity analyses were also conducted for each gene in order to compare the results between evolutionary history and sequence similarity. Phylogenetic trees were then done with either one gene or multiple genes to confirm results.

# Materials and Methods
## Genomes and Gene Sets
Publicly available full genomes were retrieved from NCBI comprising of ten type strains from *Mixta*, *Pantoea*, *Erwinia*, *Tatumella*, *Enterobacter*, and *Pseudomonas* (Table 1). *Enterobacter cloacae* is in the same order of *Enterobacterales* as *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella* whereas *Pseudomonas syringae* is in the same class of *Gammaproteobacteria*. The genomes were annotated with Prokka v1.14.1 [@Seemann2014] using default parameters, and homologous genes were extracted using GET_HOMOLOGUES software package [@Contreras-Moreira2013] with the bidirectional best-hit search algorithm using default parameters. GET_HOMOLOGUES extracted 954 homologous genes between the ten species.

@@

## Genetic Distance
### Closest Relative from Genetic Distances
Genetic distance gives a numeric estimate of evolutionary change between species. The species that has the shortest genetic distance to a species of interest is considered to be closest relative. For the following analysis, model testing was done first to estimate the best model for each gene. Next, individual phylogenetic gene trees were created and it was from these that distance matrices were exported. *P. septica* was the closest relative to 66.3% and 66.5% of the *M. calida* and *M. gaviniae* genes, respectively, for the nucleotide sequences (Figure 1). In comparison, only 42.7% and 42.8% of *M. calida* and *M. gaviniae* genes, respectively, are closely related to *P. septica* for the amino acid sequences (Figure 1), but this is still the majority. These results suggest that *P. septica* is the closest relative to *Mixta* followed by *P. agglomerans*. For the amino acid sequences, instead of the *Mixta* genes being more closely related to *P. septica*, they are more closely related to the other species in this study (Figure 1). For example, the percentage of genes related to *P. agglomerans* increased from 13.9% for both of the *Mixta* species to 22.3 and 22.4% to *M. calida* and *M. gaviniae*, respectively. This may be due to a loss of phylogenetic signal when using amino acid sequences as a result of codon degeneracy and having a third of available sites.

@@

If genetic distance infers the same closest relative for a *Mixta* gene for both nucleotide and amino acid analyses, then these results may be more reliable. Thus, for genes where the closest relative is an *Erwinia*, *Tatumella*, or *Enterobacter* species both nucleotide and amino acid analyses, they may have resulted from a horizontal gene transfer (HGT) event rather than vertical evolution from a *Pantoea* species. No *Mixta* genes were found to be closely related to *P. syringae* for either the nucleotide or amino acid sequences (Figure 1). This confirms that *P. syringae* is a suitable outgroup for this analysis and suggests that any HGT events may be restricted to the order *Enterobacterales*.

@@

```{r Figure3, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 3. Closest relatives to *M. calida* genes for nucleotide sequences around the *M. calida* genome. Length and colour of the bars represent the different species to which the gene is most closely related according to genetic distance. The *M. calida* genome has 4084 genes. No distinct patterns emerged; however, genes that are most closely related to *P. septca* are the most common and occur throughout the entire genome. Therefore, *P. septica* and *Mixta* are likely to have a common recent ancestor.", fig.width=6.5, fig.height=5}
FR_nucl_calida <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure3_Circular_NT.csv", stringsAsFactors = FALSE) %>%
  mutate(Closest_Relative = factor(Closest_Relative, 
                                   levels = c("P_septica", "P_agglomerans", "E_tasmaniensis", "E_amylovora", "T_ptyseos", 
                                              "T_saanichensis", "E_cloacae", "P_syringae"), ordered = TRUE))

ggplot(FR_nucl_calida, aes(x = ID, y = Relative_Number, fill = Closest_Relative)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 15) +
  coord_polar() +
  scale_fill_brewer(palette = "Paired", 
                    labels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", "E. cloacae",
                               "P. syringae")) +
  scale_y_continuous(limits = c(0, 9), breaks = c(0, 2, 4, 6, 8)) +
  labs(fill = "Closest Relative") +
  theme_bw() +
  theme(legend.text = element_text(face = "italic"),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(size = 12,  family = "Times New Roman"))
```

@@

```{r Figure4, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 4. Closest relatives to *M. calida* genes for amino acid sequences around the *M. calida* genome. Length and colour of the bars represent the different species to which the gene is most closely related according to genetic distance. The *M. calida* genome has 4084 genes. No distinct patterns emerged; however, genes that are most closely related to *P. septca* are the most common and occur throughout the entire genome. Therefore, *P. septica* and *Mixta* are likely to have a common recent ancestor.", fig.width=6.5, fig.height=5}
FR_aa_calida <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure4_Circular_AA.csv", stringsAsFactors = FALSE) %>%
  mutate(Closest_Relative = factor(Closest_Relative, 
                                   levels = c("P_septica", "P_agglomerans", "E_tasmaniensis", "E_amylovora", "T_ptyseos", 
                                              "T_saanichensis", "E_cloacae", "P_syringae"), ordered = TRUE))

ggplot(FR_aa_calida, aes(x = ID, y = Relative_Number, fill = Closest_Relative)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 15) +
  coord_polar() +
  scale_fill_brewer(palette = "Paired", 
                    labels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", "E. cloacae",
                               "P. syringae")) +
  scale_y_continuous(limits = c(0, 9), breaks = c(0, 2, 4, 6, 8)) +
  labs(fill = "Closest Relative") +
  theme_bw() +
  theme(legend.text = element_text(face = "italic"),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(size = 12,  family = "Times New Roman"))
```

@@

## Nucleotide and Amino Acid Identity
### Most Similar Taxa from Site Identity
Site identity is here defined as the amount of similarity between two sequences. If two sequences have a site identity of 100%, then they are identical. If site identity is 99%, then there is up to a few bases or amino acids that are different between the two sequences depending on the length of the sequences. Within each gene file, the ten sequences were aligned to both *M. calida* and *M. gaviniae* in turn and the site identity was calculated. Site identity was divided by the length of aligned genes to give a percent identity that could be compared across the genes and between the nucleotide and amino acid sequences. From these values, the closest non-*Mixta* relative was extracted for each gene.

The majority of the *Mixta* sequences were most identical with *P. septica* for nucleotide sequences (Figure 5) with 80.6 and 79.3% of *M. calida* and *M. gaviniae* sequences, respectively. This is an increase of approximately 15% from the genetic distance results for the closest relative (Figure 1). The rest of the genes are most similar to *P. agglomerans*, followed by the *Erwinia* species, *E. cloacae*, and then the *Tatumella* species. Since mutations that do not disrupt the bacterium's ability to thrive and reproduce do not occur frequently and yet may occur at constant rate, then these results suggest that *P. septica* has the most recent common ancestor with the *Mixta* species.

```{r Figure5, echo = FALSE, message = FALSE, warning=FALSE, fig.cap="Figure 5. The percentage of genes (out of 799) that were most closely related to any of the eight non-*Mixta* species using percent site identities. The species with the highest percent identity to each of the *Mixta* species is considered to be most similar taxon for a single gene. Blue bars are the results for the nucleotide sequences where the light blue bars represent *M. calida* sequences and the dark blue bars represent *M. gaviniae* sequences. Green bars are the results for the amino acid sequences where the light green bars represent *M. calida* sequences and the dark green bars represent *M. gaviniae* sequences.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure5_FirstRelativeMixtaNTAAIdentity.csv", 
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", 
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("calida_NT", "gaviniae_NT", "calida_AA", "gaviniae_AA"), ordered = TRUE))

In contrast, only 47.8 and 47.6% of *M. calida* and *M. gaviniae* sequences were most similar to *P. septica* for the amino acid sequences (Figure 5). Instead, the genes are most similar to either *P. agglomerans* or one of the two *Erwinia* species (Figure 5). Additionally, 3.1 and 2.5% more genes were most similar to the *E. cloacae* sequences for *M. calida* and *M. gaviniae*, respectively (Figure 5). In the identity analyses, amino acids that are functionally equivalent were not considered. This may explain the large differences between the nucleotide and amino acid identities for which species was most similar to the *Mixta* sequences. Additionally, at every site for a nucleotide sequence, each base has a 1:3 ratio of being identical to the base at the corresponding location of the other sequence whereas there is only a 1:19 ratio of an amino acid being identical to the amino acid at the same position of the other sequence. Thus, it is far more likely for a nucleotide at a random position to be identical the nucleotide at the same position of the other sequence. This may lead to problems when only amino acid sequences are considered, especially if functional equivalence is not considered.

Although the numbers of genes that are most similar to *P. septica* are similar between the two *Mixta* species for both nucleotide and amino acid sequences (Figure 5), there were many genes in which the same species was the not the most similar taxa to both *Mixta* species. There were 86 genes for the nucleotide sequences and 147 genes for the amino acid sequences. However, of the 86 genes for nucleotide sequences, it was the case for 52% that one of the *Mixta* sequences was most similar to one representative of a genus and the other *Mixta* sequences was most similar to the other species of the same genus. This was also the case for 62% of the 147 amino acid genes.

Furthermore, percent identity analyses give only how similar are two sequences. They do not use evolutionary models or penalize transitions and transversions in nucleotide sequences. However, because mutations that do not disrupt any vital processes are rare, percent identity analyses may lend support to relatedness between two taxa.

### Spatial Patterns of Most Similar Taxa
Similar to what was done with the closest relatives in the genetic distance analysis, spatial patterns of the most similar taxa to the *M. calida* sequences based on site identity were investigated for the nucleotide sequences (Figure 6) and the amino acid sequences (Figure 7). Once again, no discernible patterns arose when spatial position on the *M. calida* genome are considered when either the nucleotide or amino acid sequences are analyzed (Figure 6 and 7). However, majority of the nucleotide genes that were most similar to the two *Tatumella* species and to *E. amylovora* occur in the first half of the genome (Figure 6).
